\texttt{ensmallen}: a flexible C++ library for efficient function optimization

### 2018-10-15 doc/mlsys_paper/paper.tex
clarifications and sharpening

\vspace*{-0.3em}
\section{Example: Learning Linear Regression Models}
\vspace*{-0.5em}

@@

%% CS: "in development" is ambiguous here;
%% CS: is it in development by AI/ML researchers in general,
%% CS: or by ensmallen developers/contributors?
% In addition, many optimization algorithms are currently in development and will
% be released in the future.

%% CS: less ambiguous proclamation that has similar effect:
We aim to expand {\tt ensmallen} with further optimization techniques
as the need arises.  Since {\tt ensmallen} is open source,
external contributions to the codebase are welcome.

@@

\vspace*{-0.3em}
\section{Automatic generation of methods}
\vspace*{-0.5em}

It is often the case that when optimizing a function $f(x)$, the computation of
the value $f(x)$ and its derivative $f'(x)$ involve the computation of identical
intermediate results.  Consider the linear regression objective function from
earlier:
%\begin{equation}
$f(\theta) = \| y - X\theta \|_F^2.$
%\end{equation}

For this objective function, the derivative $f'(x)$ is the similar $-2 X^T (y -
X \theta)$, and both depend on the computation of the vector term $(y - X
\theta)$.  If $X$ is a large matrix, then this may be computationally expensive
to compute.  Existing optimization frameworks do not have an easy way to avoid
this duplicate computation; however, in many cases, an optimization algorithm
may need the values of both $f(\theta)$ and $f'(\theta)$ for a given $\theta$.

Using template metaprogramming, {\tt ensmallen} provides an easy (and
optional) way for users to avoid this extra computational overhead.  Instead of
specifying individual {\tt Evaluate()} and {\tt Gradient()} functions, a user
may simply write an {\tt EvaluateWithGradient()} function that returns both the
objective value and the gradient value for an input $\theta$.  For our earlier
\texttt{LinearRegressionFunction}, we could use an implementation that only
computes $(y - X \theta)$ once:

\vspace*{-0.5em}
\begin{minted}[fontsize=\small]{c++}
    double EvaluateWithGradient(const arma::mat& theta, arma::mat& gradient) {
      const arma::rowvec v = (y - X * theta); // Cache result.
      gradient = -2 * X.t() * v;
      return arma::accu(v % v); // Take squared norm of v.
    }
\end{minted}
\vspace*{-0.5em}

@@

This method could be provided {\it instead} of {\tt Evaluate()} and {\tt
Gradient()}.  In either case, template metaprogramming techniques are used to
detect which methods exist, and a wrapper class will use suitable mix-ins in
order to provide any `missing` functionality~\cite{smaragdakis2000mixin}.  For
instance, if {\tt EvaluateWithGradient()} is not provided, a version will be
generated that calls both {\tt Gradient()} and {\tt Evaluate()} in turn.
Similarly, if {\tt Evaluate()} or {\tt Gradient()} does not exist, then {\tt
EvaluateWithGradient()} is called, and the unnecessary part of the result will
be discarded.

Since all of this code generation is done at compile-time and not
at runtime, the compiler is able to perform dead code elimination
on the unused result, and may be able to avoid calculations entirely.  In some
cases, then, the generated {\tt Evaluate()} or {\tt Gradient()} functions may be
equivalently fast to what would be hand-written!

Overall, this code generation functionality reduces the requirements for users
when they are implementing their own objective functions to be optimized.  At
the time of this writing, this is implemented the most commonly-used cases:
full-batch and small-batch {\tt Evaluate()}, {\tt Gradient()}, and {\tt
EvaluateWithGradient()}.  In the future, this support may be expanded to other
sets of methods for other types of objective functions.

>>>>

\vspace*{-0.3em}
\section{Example: Learning Linear Regression Models}
\label{sec:linreg_example}
\vspace*{-0.5em}

@@

In addition, many optimization algorithms are currently in development and will
be released in the future.

@@

\vspace*{-0.3em}
\section{Automatic Generation of Methods}
\vspace*{-0.5em}

When optimizing a given function $f(x)$, the computation of
the objective function $f(x)$ and its derivative $f'(x)$ often involve the computation of identical
intermediate results.  Consider the linear regression objective function described
in Section~\ref{sec:linreg_example}, $f(\theta) = \| y - X\theta \|_F^2$.
For this objective function, the derivative $f'(x)$ has a related form of
$-2 X^T (y -X \theta)$.  Both the objective function and the derivative 
depend on the computation of the vector term $(y - X \theta)$,
which can be computationally expensive to compute if $X$ is a large matrix.
Existing optimization frameworks do not have an easy way to avoid
this duplicate computation. In many cases, an optimization algorithm
may need the values of both $f(\theta)$ and $f'(\theta)$ for a given $\theta$.

Using template metaprogramming, {\tt ensmallen} provides an easy (and
optional) way for users to avoid this extra computational overhead.  Instead of
specifying individual {\tt Evaluate()} and {\tt Gradient()} functions, a user
may simply write an {\tt EvaluateWithGradient()} function that returns both the
objective value and the gradient value for an input $\theta$.  For the 
\texttt{LinearRegressionFunction} in Section~\ref{sec:linreg_example},
we can use an implementation that only computes $(y - X \theta)$ once:

\vspace*{-0.5em}
\begin{minted}[fontsize=\small]{c++}
    double EvaluateWithGradient(const arma::mat& theta, arma::mat& gradient) {
      const arma::rowvec v = (y - X * theta); // Cache result
      gradient = -2 * X.t() * v;
      return arma::accu(v % v); // Take squared norm of v
    }
\end{minted}
\vspace*{-0.5em}

@@

This method can be provided {\it instead} of {\tt Evaluate()} and {\tt
Gradient()}.  In either case, template metaprogramming techniques are used to
detect which methods exist, and a wrapper class will use suitable mix-ins in
order to provide any `missing` functionality~\cite{smaragdakis2000mixin}.  For
instance, if {\tt EvaluateWithGradient()} is not provided, a version will be
automatically generated that calls both {\tt Gradient()} and {\tt Evaluate()} in turn.
Similarly, if {\tt Evaluate()} or {\tt Gradient()} does not exist, then {\tt
EvaluateWithGradient()} is called, and the unnecessary part of the result will
be discarded.

Since all of this code generation is automatically done at compile-time and not
at runtime, the compiler is able to perform dead code elimination
on the unused result, and may be able to avoid unnecessary calculations.  In some
cases, the generated {\tt Evaluate()} or {\tt Gradient()} functions can be
as fast as their corresponding hand-written versions.

%% CS: a subset of the text from the problematic "Templates for speed" section above
%% CS: can be placed here

Overall, the automatic code generation functionality in {\tt ensmallen}
reduces the requirements for users
when they are implementing their own objective functions to be optimized.
This leads to both quicker development and reduces the likelihood of bugs.

At the time of writing, the automatic code generation
is implemented for the most commonly-used cases:
full-batch and small-batch {\tt Evaluate()}, {\tt Gradient()}, and {\tt
EvaluateWithGradient()}.  We aim to expand this support to other
sets of methods for other types of objective functions.
